---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

# Phyloseq objects

## 2020-09-12 Simplified phyloseq object for bias analysis

The phyloseq object stored in "2020-09-12-phyloseq-focal-strains.Rds" was created by "code/2020-09-12-import-to-phyloseq.R". It contains a phyloseq object with the current abundance data from A1, S1, and S2 for a simplified feature set corresponding to just the main experimental strains: the inoculum strains and the _T. mobilis_ control.

```{r}
library(here)
library(speedyseq)

ps <- readRDS(here("output/phyloseq", "2020-09-12-phyloseq-focal-strains.Rds"))
ps
```

The taxonomy table contains the GTDB taxonomy and an additional column, `strain_group`, that can be used to subset to just the inoculum strains,
```{r}
ps0 <- ps %>% 
  subset_taxa(strain_group == "Inoculum")
```

Each feature (taxon) corresponds to a strain; however, the feature "Escherichia coli" will also include reads from the second _Escherichia coli_ strain that colonized the mice. _Faecalibacterium prausnitzii_ is included, though it did not colonize the mice. All samples are included; the `specimen_type` variable can be used to exclude the controls. Therefore, prior to analyzing bias one may wish to subset as follows,
```{r}
ps1 <- ps %>% 
  subset_samples(specimen_type %in% c("Fecal", "Inoculum")) %>%
  subset_taxa(
    strain_group == "Inoculum" & 
    genus != "Faecalibacterium" &
    genus != "Escherichia" 
  )
sample_data(ps1)$specimen_type %>% table
taxa_names(ps1)
```



