
This doc corrects a mistake in "process-maps.Rmd" that resulted in the host read counts being left out of the final count matrix; see https://github.com/mikemc/vivo-vitro/issues/8 for more info.
It loads the bacterial counts in "s{1,2}-counts.Rds" and the host counts in the BBMAP output, then combines these into a matrix.
The procedure is taken exactly from L165-L250 of "process-maps.Rmd" except for the fix in the final step to include the host reads.


```{r}
library(here)
library(fs)
library(tidyverse)
```

Map from 

```{r}
well_map <- here("output/sample-data", "dna-sample-data.Rds") %>% 
  readRDS %>%
  transmute(
    dna_sample_id,
    # Get well in "A01" format
    well = str_glue('{row}{str_pad(column, 2, "left", "0")}')
  ) %>%
  glimpse
```

```{r}
s1 <- readRDS(here("analysis", "2020-10-07-burst-pipeline", "output",
    "s1-counts.Rds"))
s2 <- readRDS(here("analysis", "2020-10-07-burst-pipeline", "output",
    "s2-counts.Rds")) %>%
  left_join(well_map, by = "well") %>%
  select(-well)
res <- bind_rows(s1, s2) %>%
  mutate(
    across(center_id, str_to_upper),
    sample_id = str_c(center_id, dna_sample_id, sep = "_")
  ) %>%
  select(sample_id, target, unambiguous_maps)
```

Get the host reads from the V1 pipeline output,
```{r}
host <- tibble(center_id = c("S1", "S2")) %>%
  mutate(
    refstats_dir = here("analysis/2020-07-09-bbmap-pipeline",
      str_to_lower(center_id), "output"),
    file = map(refstats_dir, dir_ls, glob = "*-host-refstats.tsv"),
  ) %>%
  unnest(file) %>%
  mutate(
    id = path_file(file) %>% str_extract("^[^\\-]+"),
    data = map(file, read_tsv, col_types = "cddddiiii")
    ) %>%
  select(-refstats_dir, -file) %>%
  unnest(data) %>%
  janitor::clean_names() %>%
  rename(assembly_name = number_name)
```

Confirm that the S2 read counts are even (since they were mapped as pairs),
```{r}
x <- host %>% 
  filter(center_id == "S2") %>% 
  pull(unambiguous_reads)
stopifnot(all(0 == x %% 2))
```

Munge and join with bacterial counts,
```{r}
host0 <- host %>%
  left_join(well_map, by = c(id = "well")) %>%
  transmute(center_id, id, 
    target = str_extract(assembly_name, "GC[AF]_[0-9]+.[0-9]+"),
    unambiguous_maps = case_when(
      center_id == "S1" ~ unambiguous_reads,
      center_id == "S2" ~ unambiguous_reads %/% 2L
    ),
    dna_sample_id = case_when(
      center_id == "S1" ~ id,
      center_id == "S2" ~ dna_sample_id
    ),
    sample_id = str_c(center_id, dna_sample_id, sep = "_")
  ) %>%
  select(sample_id, target, unambiguous_maps)
res_with_host <- bind_rows(res, host0)
```

## Create count matrix

```{r}
mat <- res_with_host %>%
  # Sort samples by center, specimen, aliquot
  separate(sample_id, c("center_id", "specimen_id", "aliquot_number"), 
    remove = FALSE, sep = "_") %>%
  mutate(across(c("specimen_id", "aliquot_number"), as.integer)) %>%
  arrange(center_id, specimen_id, aliquot_number) %>%
  # pivot into wide form, then convert to matrix
  select(sample_id, target, unambiguous_maps) %>%
  pivot_wider(names_from = target, values_from = unambiguous_maps,
    names_sort = TRUE, values_fill = 0) %>%
  column_to_rownames("sample_id") %>%
  as("matrix")
```

```{r}
colnames(mat)
#>  [1] "GCF_000001635.26" "GCF_000011065.1"  "GCF_000017765.1" 
#>  [4] "GCF_000020225.1"  "GCF_000020605.1"  "GCF_000022565.1" 
#>  [7] "GCF_000144325.1"  "GCF_000154125.1"  "GCF_000154205.1" 
#> [10] "GCF_000156535.1"  "GCF_000162015.1"  "GCF_000169015.1" 
#> [13] "GCF_000169035.1"  "GCF_000173815.1"  "GCF_000211215.1" 
#> [16] "GCF_000296465.1"  "GCF_000327045.1"  "GCF_000378005.1" 
#> [19] "GCF_000409585.1"  "GCF_000466485.1"  "GCF_000503755.1" 
#> [22] "GCF_001318345.1"  "GCF_001547735.1"  "GCF_001870225.1" 
#> [25] "GCF_003546975.1"  "GCF_003970495.1" 
```

The host is now included (the first OTU).

```{r}
saveRDS(mat, here("output/shotgun/v2", "2021-06-08-count-matrix.Rds"))
```

## Session info

```{r}
sessioninfo::session_info()
#> ─ Session info ─────────────────────────────────────────────────────────
#>  setting  value                       
#>  version  R version 4.1.0 (2021-05-18)
#>  os       Arch Linux                  
#>  system   x86_64, linux-gnu           
#>  ui       X11                         
#>  language (EN)                        
#>  collate  en_US.UTF-8                 
#>  ctype    en_US.UTF-8                 
#>  tz       America/New_York            
#>  date     2021-06-08                  
#> 
#> ─ Packages ─────────────────────────────────────────────────────────────
#>  package     * version date       lib source        
#>  assertthat    0.2.1   2019-03-21 [1] CRAN (R 4.0.0)
#>  backports     1.2.1   2020-12-09 [1] CRAN (R 4.0.3)
#>  broom         0.7.6   2021-04-05 [1] CRAN (R 4.0.5)
#>  cellranger    1.1.0   2016-07-27 [1] CRAN (R 4.0.0)
#>  cli           2.5.0   2021-04-26 [1] CRAN (R 4.1.0)
#>  colorspace    2.0-1   2021-05-04 [1] CRAN (R 4.1.0)
#>  crayon        1.4.1   2021-02-08 [1] CRAN (R 4.0.4)
#>  DBI           1.1.1   2021-01-15 [1] CRAN (R 4.0.4)
#>  dbplyr        2.1.1   2021-04-06 [1] CRAN (R 4.0.5)
#>  dplyr       * 1.0.6   2021-05-05 [1] CRAN (R 4.1.0)
#>  ellipsis      0.3.2   2021-04-29 [1] CRAN (R 4.1.0)
#>  fansi         0.5.0   2021-05-25 [1] CRAN (R 4.1.0)
#>  forcats     * 0.5.1   2021-01-27 [1] CRAN (R 4.0.4)
#>  fs          * 1.5.0   2020-07-31 [1] CRAN (R 4.0.2)
#>  generics      0.1.0   2020-10-31 [1] CRAN (R 4.0.3)
#>  ggplot2     * 3.3.3   2020-12-30 [1] CRAN (R 4.0.4)
#>  glue          1.4.2   2020-08-27 [1] CRAN (R 4.0.2)
#>  gtable        0.3.0   2019-03-25 [1] CRAN (R 4.0.0)
#>  haven         2.4.1   2021-04-23 [1] CRAN (R 4.1.0)
#>  here        * 1.0.1   2020-12-13 [1] CRAN (R 4.0.5)
#>  hms           1.1.0   2021-05-17 [1] CRAN (R 4.1.0)
#>  httr          1.4.2   2020-07-20 [1] CRAN (R 4.0.2)
#>  janitor       2.1.0   2021-01-05 [1] CRAN (R 4.0.4)
#>  jsonlite      1.7.2   2020-12-09 [1] CRAN (R 4.0.3)
#>  lifecycle     1.0.0   2021-02-15 [1] CRAN (R 4.0.4)
#>  lubridate     1.7.10  2021-02-26 [1] CRAN (R 4.0.4)
#>  magrittr      2.0.1   2020-11-17 [1] CRAN (R 4.0.3)
#>  modelr        0.1.8   2020-05-19 [1] CRAN (R 4.0.0)
#>  munsell       0.5.0   2018-06-12 [1] CRAN (R 4.0.0)
#>  nvimcom     * 0.9-102 2021-05-20 [1] local         
#>  pillar        1.6.1   2021-05-16 [1] CRAN (R 4.1.0)
#>  pkgconfig     2.0.3   2019-09-22 [1] CRAN (R 4.0.0)
#>  ps            1.6.0   2021-02-28 [1] CRAN (R 4.0.4)
#>  purrr       * 0.3.4   2020-04-17 [1] CRAN (R 4.0.0)
#>  R6            2.5.0   2020-10-28 [1] CRAN (R 4.0.3)
#>  Rcpp          1.0.6   2021-01-15 [1] CRAN (R 4.0.3)
#>  readr       * 1.4.0   2020-10-05 [1] CRAN (R 4.0.3)
#>  readxl        1.3.1   2019-03-13 [1] CRAN (R 4.0.0)
#>  reprex        2.0.0   2021-04-02 [1] CRAN (R 4.0.5)
#>  rlang         0.4.11  2021-04-30 [1] CRAN (R 4.1.0)
#>  rprojroot     2.0.2   2020-11-15 [1] CRAN (R 4.0.3)
#>  rstudioapi    0.13    2020-11-12 [1] CRAN (R 4.0.3)
#>  rvest         1.0.0   2021-03-09 [1] CRAN (R 4.0.5)
#>  scales        1.1.1   2020-05-11 [1] CRAN (R 4.0.0)
#>  sessioninfo   1.1.1   2018-11-05 [1] CRAN (R 4.0.0)
#>  snakecase     0.11.0  2019-05-25 [1] CRAN (R 4.0.0)
#>  stringi       1.6.2   2021-05-17 [1] CRAN (R 4.1.0)
#>  stringr     * 1.4.0   2019-02-10 [1] CRAN (R 4.0.0)
#>  tibble      * 3.1.2   2021-05-16 [1] CRAN (R 4.1.0)
#>  tidyr       * 1.1.3   2021-03-03 [1] CRAN (R 4.0.4)
#>  tidyselect    1.1.1   2021-04-30 [1] CRAN (R 4.1.0)
#>  tidyverse   * 1.3.1   2021-04-15 [1] CRAN (R 4.1.0)
#>  utf8          1.2.1   2021-03-12 [1] CRAN (R 4.0.5)
#>  vctrs         0.3.8   2021-04-29 [1] CRAN (R 4.1.0)
#>  withr         2.4.2   2021-04-18 [1] CRAN (R 4.0.5)
#>  xml2          1.3.2   2020-04-23 [1] CRAN (R 4.0.0)
#> 
#> [1] /home/michael/R/x86_64-pc-linux-gnu-library/4.0
#> [2] /usr/lib/R/library
```
