---
author: Michael McLaren
---

```{r}
library(tidyverse)
library(here)
```

This script tidies up the lab colonization assay results from the file
"colonization-summary.xlsx". The original source is the file
"Summary_Colonization.docx" received via email from Angie Mordant on
2020-02-25. To obtain "colonization-summary.xlsx", I copied this table into a
[Google Sheet](https://docs.google.com/spreadsheets/d/1w26gw4avdt8UYobSr1JW7w974VPFmzR08Ua2o9zHibQ/),
filled in all green cells with "TRUE", fixed some spelling errors, and edited
the column titles, then exported to an Excel file.

```{r}
fn <- here("data/lab", "colonization-summary.xlsx")
tb <- c(female = 1, male = 2) %>%
  map_dfr(~readxl::read_excel(fn, sheet = .), .id = "host_sex")
glimpse(tb)
```

```{r}
tb1 <- tb %>%
  rename_at("Species", str_to_lower) %>%
  pivot_longer(c(-host_sex, -species), 
    names_to = "week_condition",
    values_to = "detected"
  ) %>%
  mutate_at("detected", replace_na, FALSE) %>%
  separate(week_condition, into = c("week", "condition"), 
    sep = "(?<=\\d)\\s+") %>%
  glimpse
write_csv(tb1, here("output/lab", "colonization-test-results.csv"))
```

