---
title: "Assign taxonomy to v2 ASVs"
author: "Michael McLaren"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
---


Load the libraries
```{r}
library(here)
library(tidyverse)
library(fs)
# library(dada2); packageVersion("dada2")

import::from(Biostrings, DNAStringSet)
```

Load the seqtabs and ASVs post chimera filtering.
```{r}
st <- c(
  A1 = here("output", "a1", "dada2", "seqtab-nochim-1.Rds"),
  A2 = here("output", "a2", "dada2", "seqtab-nochim-1.Rds")
) %>%
  map(readRDS)
asvs <- st %>% 
  map(colnames) %>%
  map(DNAStringSet) %>%
  {c(.[[1]], .[[2]])}
```

## IDTAXA

### GTDB 95 using IDTAXA

Database downloaded from http://www2.decipher.codes/Downloads.html on
2020-11-19.

```{r}
packageVersion("DECIPHER")
#> [1] ‘2.18.1’
import::from(DECIPHER, IdTaxa)

# Load GTDB training set into var `trainingSet`
load(here("data/16s-taxonomy-databases/idtaxa/GTDB_r95-mod_August2020.RData"))
```

```{r}
dna <- st %>% 
  map(colnames) %>%
  map(DNAStringSet) %>%
  {c(.[[1]], .[[2]])}
```

Assign with the IDTaxa algorithm,
```{r}
ids <- IdTaxa(dna, trainingSet, threshold = 50, 
  strand = "top", processors = 3, verbose = TRUE)
saveRDS(ids, here("output/amplicon", "taxonomy", "idtaxa-gtdb-v95.Rds"))
```

```{r}
ids
ids[[1]]
ids[[2]]
```

Strange that the second ASV failed to be assigned below domain.
```{r}
dna[[1]]
dna[[2]]
```

### Silva 138 using IDTAXA

Database downloaded from http://www2.decipher.codes/Downloads.html on
2020-02-19.

```{r}
# Load Silva training set into var `trainingSet`
load(here("data/16s-taxonomy-databases/idtaxa/SILVA_SSU_r138_2019.RData"))
```

Assign with the IDTaxa algorithm,
```{r}
ids.silva <- IdTaxa(dna, trainingSet, threshold = 50, 
  strand = "top", processors = 3, verbose = TRUE)
saveRDS(ids.silva, here("output/amplicon", "taxonomy", "idtaxa-silva-v138.Rds"))
```

```{r}
ids.silva
ids.silva[[1]]
ids.silva[[2]]
```

This is E. coli! Not a great sign that IdTaxa+GTDB can't classify E. coli below
domain.

## DADA2

### Silva 138 using the DADA2-RDP algorithm

Databases downloaded from https://zenodo.org/record/3986799. Note, this is an
updated version from what was used in the pilot.

```{r silva138}
tax.dada2.silva <- dada2::assignTaxonomy(dna,
  here("data/16s-taxonomy-databases/dada2/silva-138/silva_nr99_v138_train_set.fa.gz"),
  multithread = TRUE, 
  minBoot = 50, outputBootstraps = TRUE,
  taxLevels = c("domain", "phylum", "class", "order", "family", "genus", "species")
)
saveRDS(tax.dada2.silva, here("output/amplicon", "dada2-silva-v138.Rds"))
tax.dada2.silva.with_species <- dada2::addSpecies(tax.dada2.silva$tax,
  here("data/16s-taxonomy-databases/dada2/silva-138",
    "silva_species_assignment_v138.fa.gz"),
  allowMultiple = TRUE
)
colnames(tax.dada2.silva.with_species) <- str_to_lower(
  colnames(tax.dada2.silva.with_species)
)
saveRDS(tax.dada2.silva.with_species, here("output/amplicon", "taxonomy",
    "dada2-silva-v138-with-species.Rds"))
```
Note, the version w/o species is a list with a tax matrix and a
bootstrap-support matrix.

```{r}
tax.dada2.silva.with_species %>%
  head(20) %>%
  as_tibble
```

