---
title: "Pairwise-bias figures for sequencing centers A1, S1, and S2"
author: Michael McLaren
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
geometry:
  - paperwidth=16in
  - paperheight=17in
  - top=1in
  - bottom=1in
  - left=1in
  - right=1in
fontsize: 12pt
header-includes:
  - \usepackage{multicol}
# for kableExtra
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
---


```{r setup, include=FALSE}
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
# Global chunk options
knitr::opts_chunk$set(
  cache = TRUE, autodep = TRUE, cache.comments = FALSE,
  include = FALSE, echo = FALSE,
  warning = TRUE, message = FALSE, 
  fig.width = 7, fig.height = 7,
  fig.align = "center"
)
```

<!-- Setup -->

```{r}
library(here)
library(tidyverse)
library(speedyseq)

library(kableExtra)

library(cowplot)
library(patchwork)
library(ggbeeswarm)
devtools::load_all("~/research/r-packages/metacal")

theme_set(theme_cowplot())
```

```{r}
my_psmelt <- function(physeq) {
  speedyseq::psmelt(physeq) %>%
    as_tibble %>%
    rename(taxon = OTU, sample = Sample, abundance = Abundance)
}
as_tibble.sample_data <- function(x) {
  x %>% as("data.frame") %>% as_tibble(rownames = "sample")
}
as_tibble.taxonomyTable <- function(x) {
  x %>% as("matrix") %>% as_tibble(rownames = "taxon")
}
```

```{r}
ps <- readRDS(here("output/phyloseq", "2020-09-12-phyloseq-focal-strains.Rds"))
ps
sam <- sample_data(ps) %>% as_tibble
tax <- tax_table(ps) %>% as_tibble %>% select(-strain_group)
tree <- phy_tree(ps)
```

```{r}
# Further restrict taxa (drop F. praus, T. mob, E. coli) and samples
ps0 <- ps %>%
  subset_taxa(
    strain_group == "Inoculum" & 
    genus != "Faecalibacterium" &
    genus != "Escherichia" 
  ) %>%
  # Drop controls
  subset_samples(specimen_type %in% c("Fecal", "Inoculum")) %>%
  # Drop any other poorly-seq'd sample(s)
  prune_samples(sample_sums(.) > 1e3, .)
ntaxa(ps) - ntaxa(ps0)
sample_data(ps) %>% pull(specimen_type) %>% table
sample_data(ps0) %>% pull(specimen_type) %>% table
```

<!-- Compute ratios -->

```{r}
# Compute ratios
ps0.pw <- ps0 %>%
  transform_sample_counts(~. + 1) %>%
  # transform_sample_counts(~close_elts(close_elts(.) + 1e-5)) %>%
  pairwise_ratios(margin = "taxa", filter = FALSE) %>%
  pairwise_ratios(margin = "samples", filter = FALSE) %>%
  subset_samples(specimen_id.1 == specimen_id.2)
pwtb <- ps0.pw %>%
  my_psmelt %>%
  rename(pair = taxon, bias = abundance) %>%
  separate(pair, c("taxon.1", "taxon.2"), sep = ":", remove = FALSE)
```

```{r}
#' Abbreviate taxon names from "Genus species" to "Gsp"
tax_abbrev <- function (taxa) {
    m <- str_match(taxa, "([A-Z])[a-z]+ ([a-z]{2})[a-z]*")
    ifelse(is.na(m[,1]), taxa, paste0(m[,2], m[,3]))
}

# Data frame for plotting taxa names on the diagonal
lbl <- pwtb %>%
  select(taxon.1, taxon.2) %>%
  distinct %>%
  filter(taxon.1 == taxon.2) %>%
  mutate(
    label = str_replace(taxon.1, " ", "\n"),
    across(c(taxon.1, taxon.2), factor, levels = tree$tip.label),
    across(c(taxon.1, taxon.2), fct_relabel, tax_abbrev),
  )

pwtb0 <- pwtb %>%
  filter(
    taxon.1 != taxon.2,
  ) %>%
  mutate(
    across(c(taxon.1, taxon.2), factor, levels = tree$tip.label),
    across(c(taxon.1, taxon.2), fct_relabel, tax_abbrev),
  )
```

<!-- Genome information -->

```{r}
strain_data <- here("output/strain-data", "strain-pheno-geno-tax.Rds") %>%
  readRDS %>%
  mutate(taxon = ncbi_species_display_name)
# copy numbers and genome sizes
gs <- strain_data %>%
  transmute(
    taxon,
    copy_number = ssu_count,
    genome_size = reference_genome_size,
    efficiency = copy_number / genome_size
  )
# get the bias for A1:S2
gs_bias <- crossing(taxon.1 = taxa_names(ps0), taxon.2 = taxa_names(ps0)) %>%
  filter(taxon.1 != taxon.2) %>%
  left_join(gs, by = c("taxon.1" = "taxon")) %>%
  left_join(gs, by = c("taxon.2" = "taxon"), suffix = c(".1", ".2")) %>%
  mutate(
    across(c(taxon.1, taxon.2), factor, levels = tree$tip.label),
    across(c(taxon.1, taxon.2), fct_relabel, tax_abbrev),
    gs_bias = efficiency.1 / efficiency.2
  )
# gs_bias %>% pull(gs_bias)
```


# Taxa information

```{r, include = TRUE}
tax %>%
  # filter(taxon %in% taxa_names(ps0)) %>%
  mutate(
    across(taxon, factor, levels = tree$tip.label),
  ) %>%
  arrange(desc(taxon)) %>%
  # knitr::kable() %>%
  # kable_styling(bootstrap_options = "striped", full_width = FALSE)
  kbl(booktabs = TRUE) %>%
  kable_styling(latex_options = "striped")
```

\ 
\ 

```{r, include = TRUE, fig.dim = c(6, 6)}
plot(tree)
```

\ 

T. mobilis, F. prausnitzii, and E. coli are excluded from the bias calculations
and figures.

\newpage

# About the figures

\begin{multicols}{3}
There are three "figures", each with 4 sub-figures.

\begin{itemize}
  \item Figure 1: Bias of (extraction) Protocol 1: Protocol 2
  \item Figure 2: Bias of (sequencing) Center S1: Center S2
  \item Figure 3: Bias of (sequencing) Center A1: Center S2
\end{itemize}

\begin{itemize}
  \item A. Extraction Batch 1, data points
  \item B. Extraction Batch 1, mean + bootstrap 95% CI
  \item C. Extraction Batch 2, data points
  \item D. Extraction Batch 2, mean + bootstrap 95% CI
\end{itemize}

Data points are ratios of ratios, representing the (differential) bias of a
pair of taxa and a pair of extraction or sequencing protocol; i.e., the
efficiency of one taxon relative to another, for one protocol relative to
another. Exactly how the data points are generated depends on whether
extraction or sequencing bias is being examined, and is described in the figure
subtitles.
\end{multicols}


# Figure 1: Extraction bias

```{r}
extraction_bias_plot <- function(batch, fun.data = NULL, prefix = "") {
  p <- pwtb0 %>%
    filter(
      extraction_batch.1 == batch,
      center_id.1 == center_id.2,
      extraction_protocol.1 == "1",
      extraction_protocol.2 == "2",
    ) %>%
    ggplot(aes(
        x = specimen_type.1:center_id.1,
        y = bias, 
        color = center_id.1:specimen_type.1)) +
    geom_hline(yintercept = 1, color = "grey") +
    scale_y_log10() +
    scale_color_brewer(type = "qual", palette = "Paired", guide = "none") +
    facet_grid(taxon.1 ~ taxon.2, scales = "free_y") +
    labs(x = "{specimen type}:{extraction protocol}") +
    theme(
      axis.text.x = element_text(angle = 90)
    ) +
    plot_annotation(
      title = str_glue(.sep = " ",
        "{prefix}Bias of Protocol 1:Protocol 2 in Batch {batch}"),
      subtitle = 
        str_c(sep = " ",
        "(row-taxon / col-taxon in aliquot extracted by P. 1)", "/",
        "(row-taxon / col-taxon in aliquot extracted by P. 2),",
        "for the same biological specimen")
    )
  if (is.null(fun.data)) {
    lyr <- geom_quasirandom(size = 0.5)
  } else {
    lyr <- stat_summary(fun.data = fun.data, fatten = 1.8)
  }
  # Note, must add text layer after due to continuous scale
  p + lyr + 
    geom_text(data = lbl, aes(x = 3.5, y = 1, label = label), color = "black")
}
```


```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 1, points
extraction_bias_plot("1", prefix = "Figure 1A. ")
```

```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 1, mean est
extraction_bias_plot("1", fun.data = mean_cl_boot, "Figure 1B. ")
```

```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 2, points
extraction_bias_plot("2", prefix = "Figure 1C. ")
```

```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 2, mean est
extraction_bias_plot("2", fun.data = mean_cl_boot, "Figure 1D. ")
```


# Figure 2: Sequencing bias, S1:S2

```{r}
sequencing_bias_plot <- function(center1, center2, batch, fun.data = NULL,
  prefix = "") {
  p <- pwtb %>%
    filter(
      extraction_batch.1 == batch,
      extraction_protocol.1 == extraction_protocol.2,
      aliquot_number.1 == aliquot_number.2,
      center_id.1 == center1, 
      center_id.2 == center2,
      taxon.1 != taxon.2,
    ) %>%
    mutate(
      across(c(taxon.1, taxon.2), factor, levels = tree$tip.label),
      across(c(taxon.1, taxon.2), fct_relabel, tax_abbrev),
      # across(starts_with("extraction_protocol"), fct_relabel, ~str_c("P", .)),
    ) %>%
    ggplot(aes(
        x = specimen_type.1:extraction_protocol.1,
        y = bias, 
        color = extraction_protocol.1:specimen_type.1)) +
    geom_hline(yintercept = 1, color = "grey") +
    scale_y_log10() +
    # scale_color_brewer(type = "qual", palette = palettes["specimen_type"]) +
    scale_color_brewer(type = "qual", palette = "Paired", guide = "none",
      drop = FALSE) +
    facet_grid(taxon.1 ~ taxon.2, scales = "free_y") +
    labs(x = "{specimen type}:{extraction protocol}") +
    theme(
      axis.text.x = element_text(angle = 90)
    ) +
    plot_annotation(
      title = str_glue(.sep = " ",
        "{prefix}Bias of Center {center1}:Center {center2}", 
        "in Batch {batch}"),
      subtitle = str_glue(.sep = " ",
        "(row-taxon / col-taxon measured by {center1})", "/",
        "(row-taxon / col-taxon measured by {center2}),",
        "for the same DNA sample")
    )
  if (is.null(fun.data)) {
    lyr <- geom_quasirandom(size = 0.5)
  } else {
    lyr <- stat_summary(fun.data = fun.data, fatten = 1.8)
  }
  # Note, must add text layer after due to continuous scale
  p + lyr + 
    geom_text(data = lbl, aes(x = 2.5, y = 1, label = label), color = "black")
}

```


```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 1, points
sequencing_bias_plot("S1", "S2", "1", prefix = "Figure 2A. ")
```

```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 1, mean est
sequencing_bias_plot("S1", "S2", "1", fun.data = mean_cl_boot, "Figure 2B. ")
```

```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 2, points
sequencing_bias_plot("S1", "S2", "2", prefix = "Figure 2C. ")
```

```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 2, mean est
sequencing_bias_plot("S1", "S2", "2", fun.data = mean_cl_boot, "Figure 2D. ")
```


# Figure 3: Sequencing bias, A1:S2



```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 1, points
sequencing_bias_plot("A1", "S2", "1", prefix = "Figure 3A. ") +
  geom_hline(data = gs_bias, aes(yintercept = gs_bias, color = "darkred"), alpha = 0.5)
```

```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 1, mean est
sequencing_bias_plot("A1", "S2", "1", fun.data = mean_cl_boot, "Figure 3B. ") +
  geom_hline(data = gs_bias, aes(yintercept = gs_bias, color = "darkred"), alpha = 0.5)
```

```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 2, points
sequencing_bias_plot("A1", "S2", "2", prefix = "Figure 3C. ") +
  geom_hline(data = gs_bias, aes(yintercept = gs_bias, color = "darkred"), alpha = 0.5)
```

```{r, include = TRUE, fig.dim = c(15, 15), out.dim = c(15, 15)}
# Batch 2, mean est
sequencing_bias_plot("A1", "S2", "2", fun.data = mean_cl_boot, "Figure 3D. ") +
  geom_hline(data = gs_bias, aes(yintercept = gs_bias, color = "darkred"), alpha = 0.5)
```


