---
title: Inspect DNA concentrations
author: Michael McLaren
date: "`r Sys.Date()`"
description:
output:
  html_document:
    toc: true
    fig_width: 6
    dev: "svg"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(progress = TRUE, verbose = TRUE)
# Global chunk options
knitr::opts_chunk$set(
  include = TRUE, echo = TRUE,
  warning = TRUE, message = FALSE, 
  fig.width = 7, fig.height = 7
)
```

Goal: Compare DNA concentrations by protocol and specimen type. This analyses
was conducted on 2020-03-23 and was updated on 2020-08-05 to use the new
results files and create an html report.

## Setup

```{r}
library(tidyverse)
library(here)
library(cowplot)
# library(ggridges)
library(ggbeeswarm)
# library(ggforce)
library(speedyseq)
import::from(metacal, mutate_by, clr)
theme_set(theme_cowplot(12))
```

```{r}
my_psmelt <- function(physeq) {
  speedyseq::psmelt(physeq) %>%
    as_tibble %>%
    rename(taxon = OTU, sample = Sample, abundance = Abundance)
}
as_tibble.phyloseq <- function(x) {
  x %>% my_psmelt
}
as_tibble.sample_data <- function(x) {
  x %>% as("data.frame") %>% as_tibble(rownames = "sample")
}
as_tibble.taxonomyTable <- function(x) {
  x %>% as("matrix") %>% as_tibble(rownames = "taxon")
}
as_tibble.XStringSet <- function(x) {
  x %>% as.character %>% enframe("taxon", "sequence")
}
```

```{r}
# Phyloseq object w/ the A1 sequencing data,
ps <- readRDS(here("results", "a1", "a1-phyloseq-silva-v138.Rds")) %>%
  filter_taxa2(~sum(.) >= 10)
sample_data(ps)$specimen_id <- sample_data(ps)$specimen_id %>% as.factor
# Strain assignments
assignments <- here("results", "a1", "a1-asv-strain-assignments.csv") %>%
  read_csv
```

From Chris at the MSMBL: "...we use the Zymo mock at a concentration of 2ng/uL.
The last QC concentration I did of the 2ng/uL stock it was 2.25ng/uL."

```{r}
sample_data(ps)["A1_zymo_posc", "dna_conc"] <- 2.25
sample_data(ps)["A1_water_negc", "dna_conc"] <- 
  min(sample_data(ps)$dna_conc) / 10
```

```{r}
tax <- tax_table(ps) %>% as_tibble
sam <- sample_data(ps) %>% as_tibble
```


## Inspect DNA concentrations

NOTE: Consider redo this with T. mobilis included, using the original
extraction metadata since that has the samples we didn't sequence.

```{r, fig.dim = c(7, 4)}
sam %>%
  filter(specimen_type %in% c("fecal", "inoculum", "T. mobilis")) %>%
  ggplot(aes(specimen_type, dna_conc, color = extraction_protocol)) +
  geom_boxplot() +
  geom_quasirandom(dodge.width = 0.75) +
  # scale_y_sqrt(breaks = c(0.1, 1, 5, 10, 20, 30)) +
  scale_y_log10() +
  facet_wrap(~extraction_batch,
    labeller = as_labeller(function(x) paste("Batch", x))
  ) +
  scale_color_brewer(type = "qual") +
  labs(
    x = "Specimen type", y = "DNA concentration (ng/uL)",
    color = "Extraction Protocol",
    title = "DNA concentration by protocol, specimen type, and batch"
  ) +
  theme(
    legend.position = "bottom",
    panel.spacing = unit(0.03, "npc")
  )
```

This figure indicates that 

* Protocol 1 gave higher DNA yields on fecal samples, but Protocol 2 gave
  higher yields on inoculum samples. 
* The extent of these differences varied between batches. In Batch 2, the
  difference was less among fecal samples and greater among inoculum samples.

Differences between batches are that in Batch 2 

* there was an additional freeze thaw cycle for fecal and inoculum samples
* we pooled multiple pellets for inoculum samples, which explains why the
  overall yield of inoculum samples is higher in batch 2 
* the reagents had been opened for some days

```{r}
sam %>%
  filter(specimen_type == "inoculum") %>%
  transmute(specimen_id, number_of_pellets / number_of_aliquots, 
    extraction_batch, collection_week) %>%
  distinct %>%
  arrange(extraction_batch)
```

A possible explanation of the difference in pattern between batches is that the
additional freeze-thaw cycle increased the yield more for Protocol 2 than for
Protocol 1. It may even have decreased the yield of protocol 1 as indicated by
its lower yield, though this could also be due to biologically lower host +
bacterial load in the sample and/or a change in composition.

Possible explanations for the difference between specimen types (Protocol 1
higher on fecal, Protocol 2 on inoculum) include:

1. Protocol 1 yields more host DNA than Protocol 2, which leads to higher
   Protocol 1 yield in fecal samples only. We can test this hypothesis once we
   have shotgun data and can compare host to bacterial sequencing reads.
2. The bacterial taxonomic compositions vary substantially between fecal and
   inoculum samples. It may be that Protocol 1 is better at extracting taxa
   that are abundant in the fecal samples relative to taxa that are more
   abundant in the inoculum samples and Protocol 2. We can test this hypothesis
   by quantifying bias from the taxonomic profiles if we assume that the bias
   we observe in the profiles is mainly due to extraction.

### Compare to library size

Note, I'm using the number of reads post DADA2, rather than the actual library
depth.

```{r, fig.dim = c(7,5)}
sam %>%
  mutate_at("extraction_protocol", fct_explicit_na) %>%
  ggplot(
    aes(dna_conc, reads_final, 
      color = extraction_protocol,
      shape = specimen_type
    )
  ) +
  geom_point() +
  scale_x_log10() +
  scale_color_brewer(type = "qual") +
  labs(
    y = "Number of reads", x = "DNA concentration (ng/uL)",
    color = "Extraction Protocol",
    title = "Read depth versus DNA concentration "
  ) +
  theme(
    legend.position = "bottom",
    panel.spacing = unit(0.03, "npc")
  )
```

There does not seem to be much relationship, which is to be expected given the
library normalization step.
